---
title: "Explorary Data Analysis for Vitamine B12 data"
author: Batool Almarzouq
Date: 3-4-2022
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
    math: katex
---
## Summary

The Final_Dataset contains 5009 patients of which 1798 of them are control and 3211 are prescribed metformin. All metformin-treated patients are in the Medication dataset. Both datasets were combined in the joined_dataset. The earlier clinic date is 2018-01-01 and the latest is 2019-11-21. The oldest patient is born on 1921-03-11 and the younger patient is born on 2004-02-15. The latest exam was carried out on 2022-02-14 16:06:34. Also, the patients come from 15 hospitals which are KAMC-R, NGCSC, HCSC, DIRAB, PBRC, ISKAN, FMPHC-M, Bahra, KAMC-J, NGSPC, Iskan, KAH, IABFH, PHC-HS, DmPHC.


## Install Packages

```{r message=FALSE, warning=FALSE}
library(readxl)
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(anytime)
library(DataExplorer)
library(skimr)

```

## Read the data from excel

```{r message=FALSE, warning=FALSE}
#read the complete dataset
Final_Dataset <- read_csv(here("data","Final_Dataset.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(age = gsub("year","", age)) 

# Make the age and exam results numeric  
Final_Dataset$age <-as.numeric(Final_Dataset$age)
Final_Dataset$exam_result <-as.numeric(Final_Dataset$exam_result)
Final_Dataset$patient_no <- as.character(Final_Dataset$patient_no)

#read the medications dataset
medications <- read_csv(here("data","medications.csv")) %>% 
  janitor::clean_names()

medications$patient_no <- as.character(medications$patient_no)

```

## Explore the Final Dataset

```{r echo=TRUE, message=FALSE, warning=FALSE}

# Have a look at the dimentions
dim(Final_Dataset)
# Have a look at all columns in the dataset
glimpse(Final_Dataset)
# Summarise each column
#summary(Final_Dataset)
# Summarise each column in a table
skim(Final_Dataset)


```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# generate a report from the dataset
DataExplorer::create_report(Final_Dataset)
```

## Explore the age of the sample and number of control or treatment

```{r}
summary(Final_Dataset$age)
summary(Final_Dataset$exam_result)

# Final_Dataset_control <- Final_Dataset %>% 
#   group_by(patient_no) %>% 
#   summarise(metformin_patients = sum(metformin_use == 0))

Final_Dataset_control <- Final_Dataset %>% 
  filter(metformin_use == 0) %>% 
  group_by(patient_no) %>% 
  summarise(n())

paste0("nember of patients who are control: ", nrow(Final_Dataset_control))

Final_Dataset_metforin <- Final_Dataset %>% 
  filter(metformin_use == 1) %>% 
  group_by(patient_no) %>% 
  summarise(n())

paste0("nember of patients who took metformin: ", nrow(Final_Dataset_metforin))
```


## Explore the Medication Dataset

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Have a look at all columns in the dataset
glimpse(medications)
# Have a look at the dimentions
dim(medications)
# Summarise each column
#summary(medications)
# Summarise each column in a table
skim(medications)

```


## Explore the hospitals and Medicine names

```{r}
# Hospital names:
unique(medications$hospital_name)
unique(medications$medicine_name_general)

```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# generate a report from the dataset
DataExplorer::create_report(Final_Dataset)
```

## Explore the Patients in each dataset

```{r echo=TRUE, message=FALSE, warning=FALSE}
# number of patients in the complete/final dataset
paste0("The number of patients in the complete/final dataset are ", length(unique(Final_Dataset$patient_no)))

#number of patients in the medications dataset
paste0("The number of patients in the medications dataset are ", length(unique(medications$patient_no)))
```

```{r}
patient_no_dataset <- unique(Final_Dataset$patient_no)
length(patient_no_dataset)

patient_no_medications <- unique(medications$patient_no)
length(patient_no_medications)

length(which(patient_no_dataset %in% patient_no_medications))

```

## Join both datasets

```{r}

joined_data <- dplyr::right_join(Final_Dataset, medications, by =c("patient_no", "clinic_date"))
dim(joined_data)

length(unique(joined_data$patient_no))

joined_data <- joined_data%>% 
  select(-ends_with(".y")) %>% 
  janitor::clean_names()

write_csv(joined_data, here("data", "joined.csv"))

```



