---
title: "Explorary Data Analysis for Vitamine B12 data"
author: Batool Almarzouq
Date: 3-4-2022
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
## Summary

The Final_Dataset contains 5009 patients of which 1798 of them are control and 3211 are prescribed metformin. All metformin-treated patients are in the Medication dataset. Both datasets were combined in the joined_dataset. The earlier clinic date is 2018-01-01 and the latest is 2019-11-21. The oldest patient is born on 1921-03-11 and the younger patient is born on 2004-02-15. The latest exam was carried out on 2022-02-14 16:06:34. Also, the patients come from 15 hospitals which are KAMC-R, NGCSC, HCSC, DIRAB, PBRC, ISKAN, FMPHC-M, Bahra, KAMC-J, NGSPC, Iskan, KAH, IABFH, PHC-HS, DmPHC.


## Install Packages

```{r message=FALSE, warning=FALSE}
library(readxl)
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(anytime)
library(DataExplorer)
library(skimr)
library(ggpubr)
```

## Read the data from excel

```{r message=FALSE, warning=FALSE}
#read the complete dataset
Final_Dataset <- read_csv(here("data","Final_Dataset.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(age = gsub("year","", age)) 

# Make the age and exam results numeric  
Final_Dataset$age <-as.numeric(Final_Dataset$age)
Final_Dataset$exam_result <-as.numeric(Final_Dataset$exam_result)
Final_Dataset$patient_no <- as.character(Final_Dataset$patient_no)

#read the medications dataset
medications <- read_csv(here("data","medications.csv")) %>% 
  janitor::clean_names()

medications$patient_no <- as.character(medications$patient_no)

#read B12 data and make columns chracters
B12 <- read_csv(here("data", "Final_medications_list.csv")) %>% 
  janitor::clean_names()

B12$patient_no <- as.character(B12$patient_no)

```

## Explore the Final Dataset

```{r echo=TRUE, message=FALSE, warning=FALSE}

# Have a look at the dimentions
dim(Final_Dataset)
# Have a look at all columns in the dataset
glimpse(Final_Dataset)
# Summarise each column
#summary(Final_Dataset)
# Summarise each column in a table
skim(Final_Dataset)


```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# generate a report from the dataset
DataExplorer::create_report(Final_Dataset)
```

## Explore the age of the sample and number of control or treatment

```{r}
summary(Final_Dataset$age)
summary(Final_Dataset$exam_result)

# Final_Dataset_control <- Final_Dataset %>% 
#   group_by(patient_no) %>% 
#   summarise(metformin_patients = sum(metformin_use == 0))

Final_Dataset_control <- Final_Dataset %>% 
  filter(metformin_use == 0) %>% 
  group_by(patient_no) %>% 
  summarise(n())


paste0("nember of patients who are control: ", nrow(Final_Dataset_control))

Final_Dataset_control  <-  Final_Dataset %>% 
  filter(metformin_use == 0) 

summary(Final_Dataset_control)

Final_Dataset_metforin <- Final_Dataset %>% 
  filter(metformin_use == 1) %>% 
  group_by(patient_no) %>% 
  summarise(n())


hist(Final_Dataset_control$age, breaks = 30)
hist(Final_Dataset_control$exam_result, breaks = 40)


paste0("nember of patients who took metformin: ", nrow(Final_Dataset_metforin))

Final_Dataset_metforin <- Final_Dataset %>% 
filter(metformin_use == 1)

summary(Final_Dataset_metforin)
hist(Final_Dataset_metforin$age, breaks = 30)
hist(Final_Dataset_metforin$exam_result, breaks = 50)

```


## Explore the Medication Dataset

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Have a look at all columns in the dataset
glimpse(medications)
# Have a look at the dimentions
dim(medications)
# Summarise each column
#summary(medications)
# Summarise each column in a table
skim(medications)

```


## Explore the hospitals and Medicine names

```{r}
# Hospital names:
unique(medications$hospital_name)
unique(medications$medicine_name_general)

```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# generate a report from the dataset
DataExplorer::create_report(Final_Dataset)
```

## Explore the Patients in each dataset

```{r echo=TRUE, message=FALSE, warning=FALSE}
# number of patients in the complete/final dataset
paste0("The number of patients in the complete/final dataset are ", length(unique(Final_Dataset$patient_no)))

#number of patients in the medications dataset
paste0("The number of patients in the medications dataset are ", length(unique(medications$patient_no)))
```

```{r}
patient_no_dataset <- unique(Final_Dataset$patient_no)
length(patient_no_dataset)

patient_no_medications <- unique(medications$patient_no)
length(patient_no_medications)

length(which(patient_no_dataset %in% patient_no_medications))

patient_no_b12_medications <- unique(B12$patient_no)
length(patient_no_b12_medications)

```

## Join both datasets

```{r}

## Join metformin with full dataset
joined_data_met <- dplyr::full_join(Final_Dataset, medications, by =c("patient_no", "clinic_date"))
dim(joined_data_met)

length(unique(joined_data_met$patient_no))

joined_data_met <- joined_data_met %>% 
  select(-ends_with(".y")) %>% 
  janitor::clean_names()  %>% 
  rename(d_o_b = d_o_b_x) 

write_csv(joined_data_met, here("data", "joined.csv"))
skim(joined_data_met)

## Join B12  with full dataset

joined_data_b12 <- dplyr::full_join(joined_data_met, B12, by =c("patient_no", "d_o_b" ,"clinic_date"))


## Join metformin and B12 dataset


joined_data_b12 <-  joined_data_b12 %>% 
  rename(exam_result_B12 = exam_result) %>% 
  select(-c(exam_detail_item, exam_name, gender_x)) %>% 
  distinct(patient_no, d_o_b, clinic_date, exam_result_B12, .keep_all = TRUE) %>% 
  select(-c(hospital_name_x, enterprise_number)) %>% 
  rename(B12_name = medicine_name_general.y) %>% 
  rename(prescription_date_B12 = prescription_date.y, prescription_date_metformin = prescription_date.x) %>% 
  select(-c(hospital_name)) %>% 
  filter(!is.na(gender)) %>% 
  filter(!is.na(exam_result_B12)) 


dim(joined_data_b12)

paste0("categories are ", unique(joined_data_met$metformin_use))
```

## Analysis

```{r}
joined_data_b12 <-  joined_data_b12 %>%
  mutate(category = ifelse(
    metformin_use == 0 & str_detect(B12_name, "ablet"),
    0,
        ifelse(
      metformin_use == 0 & str_detect(B12_name, "njection"),
      1,
    ifelse(
      metformin_use == 1 & str_detect(B12_name, "ablet"),
      2,
      ifelse(metformin_use == 1 & str_detect(B12_name, "njection"),
             3, 4)
    )
  )))


```


```{r}

data <- write.csv(joined_data_b12, here("data", "joined.csv"))

```


```{r}
library(reshape2)
reshaped <- joined_data_b12 %>%
  pivot_wider(names_from = category, 
              values_from = exam_result_B12, 
              values_fill = 0) %>% 
  select(patient_no,"0","1","2","3")

data <- write.csv(joined_data_b12, here("data", "reshaped.csv"))


```


```{r}
my_comparisons_met <- list( c("0", "1"), c("0", "2"),c("0", "3"), c("1", "2"), c("1", "3"), c("2", "3"))

joined_data_b12 %>% 
    mutate(category = factor(category)) %>% 
    ggboxplot(x = "category", y = "exam_result_B12",
                color = "category", palette =c("#00AFBB", "#E7B800", "#FC4E07", "#FC4E09"),
                add = "jitter", shape = "category") +
    xlab("Metformin Use") +
    ylab("B12 Level") +
    scale_x_discrete(labels=c ("- Met/B12 tablet","- Met/B12 Injection", "+ Met/B12 tablet", "+ Met/B12 Injection")) +
    stat_compare_means(comparisons =  my_comparisons_met, label = "p.signif", method = "t.test")  # Add pairwise comparisons p-value


my_comparisons_gender <- list( c("Male", "Female"))

joined_data_b12 %>% 
    mutate(gender = factor(gender)) %>% 
    ggboxplot(x = "gender", y = "exam_result_B12",
                color = "gender" , palette =c("#00AFBB", "#E7B800"),
                add = "jitter", shape = "gender") +
    xlab("Gender") +
    ylab("B12 Level") +
    stat_compare_means(comparisons =  my_comparisons_gender, label = "p.signif", method = "t.test")  # Add pairwise comparisons p-value

  
```



```{r}

model <- lm(exam_result_B12 ~ age+ gender+bmi,joined_data_b12)

summary(model)
```

